---
title: "Biomarkers"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
---
```{r setup, echo=FALSE, message=FALSE}
library(readxl)
library(tidyverse)
library(broom)
library(data.table)
library(olsrr)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```
### Import data
```{r import data}
#Pull APACHE score in from outcomes data
outcome <- read_excel("~/Documents/ROCIanalysis/ROCI_101918.xlsx", sheet = "Outcomes", range = cell_cols(1:23))
outcome <- outcome %>% dplyr::select(Subject_ID, APACHE, Final_Dx) 

#Imaging/score data
images <- read_excel("~/Documents/ROCIanalysis/ROCI_101918.xlsx", sheet = "Imaging", range = cell_cols(1:16))
images <- images %>% dplyr::select(Subject_ID, admit_cxr, Total) %>% rename(score = Total) %>%
  filter(admit_cxr == 1) %>% left_join(outcome, by="Subject_ID") %>% filter(Final_Dx != "Control") %>% mutate(quartile = ntile(score,4))

#Biomarker data
biomarker <- read_csv("~/Documents/ROCIanalysis/IL18_mtDNA.csv")
biomarker <- biomarker %>% dplyr::select(SID, Date_bld_draw, IL18D0, nadh_copies, HMITO_38, HMITO_96, bglob, Lactate) %>% 
  rename(Subject_ID = SID, blood_date = Date_bld_draw) %>%
  filter(!is.na(blood_date)) %>% mutate(blood_date = as.Date(blood_date, "%m/%d/%Y")) %>% 
  right_join(images, by = "Subject_ID") 

#log transforms
biomarker <- biomarker %>% mutate(logIL18 = log10(IL18D0), logbglob = log10(bglob), lognadh = log10(nadh_copies), logh96 = log10(HMITO_96))
```

<p>Per Circulating mitochondrial DNA in patients in the ICU as a marker of mortality: derivation and validation (PLoS Med 2013):
<br>mtDNA = human NADH dehydrogenase 1 gene (represented in this data as **nadh_copies**)
<br>nuclear DNA = human B-globin (represented in this data as **bglob**)
<br>Innate immune response (represented in this data by **IL18D0**, meaning collected wi 48H admission)
<br>

### Summary stats
```{r summary stats}
#subjects with both a scored CXR and a given biomarker
markers <- biomarker %>% dplyr::select(IL18D0, nadh_copies, HMITO_38, HMITO_96, bglob, Lactate)

#define functions to accomodate missing data
count_rows <- function(x){
  NROW(na.omit(x))
}
max_val <- function(x){
  max(na.omit(x))
}
min_val <- function(x){
  min(na.omit(x))
}

#Build the summary table
sum_stat <- data.table(IL18D0=double(), nadh_copies=double(), HMITO_38=double(), HMITO_96=double(), bglob=double(), Lactate=double())
sum_stat <- rbind(sum_stat, t(apply(markers, 2, count_rows)))
sum_stat <- rbind(sum_stat, t(apply(markers, 2, max_val)))
sum_stat <- rbind(sum_stat, t(apply(markers, 2, min_val)))
sum_stat <- cbind(sum_stat, c("n", "max", "min"))
sum_stat <- sum_stat[,c(7,1,2,3,4,5,6)]
sum_stat
```

### Visualize data
## Scatterplot
```{r scatterplot}
biomarker %>% dplyr::select(IL18D0, nadh_copies, HMITO_38, HMITO_96, bglob, Lactate, score) %>% gather(marker, value, -score, na.rm=T) %>% 
  ggplot(aes(score, value)) + geom_point() + scale_y_log10() + facet_grid(.~marker) + ylab("value (log scale)") + xlab("Total CXR score")

#To check the integrity of the gather, name the data and check against:
#vert %>% group_by(marker) %>% summarize(count = length(marker)) #check that the gather did not lose any data
```

```{r scatterplot with log transforms}
biomarker %>% dplyr::select(logIL18, lognadh, logbglob, score) %>% gather(marker, value, -score, na.rm=T) %>% 
  ggplot(aes(score, value)) + geom_point() + facet_grid(.~marker) + ylab("value (log scale)") + xlab("Total CXR score") + geom_smooth(method = "lm")
```

```{r scatter quartile}
biomarker %>% dplyr::select(logIL18, lognadh, logbglob, quartile) %>% gather(marker, value, -quartile, na.rm=T) %>% 
  ggplot(aes(quartile, value)) + geom_jitter(width=0.2) + facet_grid(.~marker) + ylab("value (log scale)") + xlab("CXR score quartile")
```


## Test for normality
```{r normality}
shapiro.test(biomarker$IL18D0)
shapiro.test(biomarker$nadh_copies)
shapiro.test(biomarker$HMITO_38)
shapiro.test(biomarker$HMITO_96)
shapiro.test(biomarker$bglob)
shapiro.test(biomarker$Lactate)
```
<p>Only HMITO_38 is normally distributed. 

###Log transformation
```{r log transform}
shapiro.test(biomarker$logIL18)
shapiro.test(biomarker$lognadh)
shapiro.test(biomarker$logh96)
shapiro.test(biomarker$logbglob)
```
<p>Log transforming IL-18 and nadh made those two normally distributed, while HMITO_96 and bglob remain non-normal

## Regressions
###Univariate
```{r univariate regression}
biomarker %>% dplyr::select(logIL18, lognadh, HMITO_38, logh96, logbglob, Lactate, score) %>% gather(marker, value, -score, na.rm=T) %>%
  group_by(marker) %>% do(tidy(lm(value~score, data = .))) %>% filter(term != "(Intercept)")

```
<p>The regression demonstrates that CXR score was a significant predictor of HMITO_38, log of nadh, log of HMITO96, log of bglob. It does not appear to be a significant predictor of lactate or IL18 (either untransformed or log transformed).

<p>Since nadh represents mtDNA and bglob represents nucDNA, I can state that CXR severity associates with those. 

###Controlling for APACHE score
```{r multivariate regression}
biomarker %>% dplyr::select(logIL18, lognadh, HMITO_38, logh96, logbglob, Lactate, score, APACHE) %>% gather(marker, value, -score, -APACHE, na.rm=T) %>%
  group_by(marker) %>% do(tidy(lm(value~score+APACHE, data = .))) %>% filter(term != "(Intercept)")
```
<p>APACHE is a significant predictor of all the biomarkers (lactate is borderline) in the model. Controlling for APACHE, CXR score is no longer predictive of bglob (log transformed, represents nucDNA) or nadh (log transformed, represents mtDNA).

##Quartile analysis

```{r quartiles}
biomarker %>% mutate(quartile = as.factor(quartile)) %>% dplyr::select(logIL18, lognadh, logbglob, Lactate, quartile) %>% gather(marker, value, -quartile, na.rm=T) %>%
  group_by(marker) %>% do(tidy(lm(value~quartile, data = .))) %>% filter(term != "(Intercept)")
```


