---
title: "Lung Weights"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
---
```{r setup, echo=FALSE, message=FALSE}
library(readxl)
library(tidyverse)
library(survival)
library(olsrr)
library(gridExtra)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```
### Loading Data
```{r load data}
#Load demographic data for gender
demo <- read_excel("~/Documents/ROCIanalysis/ROCI_101918.xlsx", sheet = "Outcomes", range = cell_cols(1:4))
demo <- demo %>% dplyr::select(-imgs, -Age)

#Load imaging/weight data and merge
weights <- read_excel("~/Documents/ROCIanalysis/ROCI_101918.xlsx", sheet = "LungWt", range = cell_cols(1:12)) %>% 
  gather(side, weight, r_lung_wt:l_lung_wt) %>%
  mutate(side = ifelse(side == "r_lung_wt", "right", "left"), side_total = ifelse(side == "right", quad1+quad3, quad2+quad4)) %>%
  filter(wi_24h == 1) %>%
  left_join(., demo, by = "Subject_ID") %>% 
  mutate(Male = as.factor(Male))

```

# Analysis

### Summary Statistics
<p> Our lungs appear to be slightly heavier than those included in the RALE sample.
```{r summary stats}
summ <- weights %>% summarise(mean = mean(weight, na.rm=T), median = median(weight, na.rm=T)) 
summ
```

```{r scatterplot}
weights %>% ggplot(aes(side_total, weight)) + geom_point() + geom_smooth(method = "lm") + xlab("CXR score") +
  ylab("Weight (g)") + ggtitle("CXR score as a predictor of lung weight at autopsy")
```


###Testing for normality
<p> The data are non-normal.
```{r normality test}
shapiro.test(weights$weight)
weights %>% ggplot(aes(weight)) + geom_histogram()
```

### Spearman
<p>Our Spearman is much lower than that reported in the RALE study. They reported rho = 0.59 for all explants and rho = 0.73 for the heavy subset. 
```{r spearman correlation}
cor.test(weights$side_total, weights$weight, method = "spearman")

rale_weights <- weights %>% filter(weight > 910)  #spearman on those greater than the median
cor.test(rale_weights$side_total, rale_weights$weight, method = "spearman")
```

### Linear regression
<p>Based on the residual diagnostics, it is reasonable to use linear regression despite the data being non-normal (below). This allows us to control for the confounders of height and gender on lung weight. 
```{r linear regression}
weights %>% ggplot(aes(side_total, weight, color = side)) + geom_point() + geom_smooth(method = "lm")

wt <- lm(weight ~ side_total+height+Male, data = weights)
summary(wt)

#check normality of errors
ols_plot_resid_qq(wt)
ols_test_correlation(wt)
```






