---
title: "Regression: Full cohort"
author: "S. Mason"
date: "10/23/2018"
output: html_document
---

```{r setup, echo=FALSE, message=FALSE}
library(readxl)
library(dplyr)
library (ggplot2)
library(survival)
library(survminer)
library(QuantPsyc)
```
```{r load data, message=FALSE, echo=FALSE}
#Load outcomes data
outcomes <- read_excel("~/Documents/ROCIanalysis/ROCI_101918.xlsx", sheet = "Outcomes", col_types = c("numeric", "skip", "numeric", "numeric", "numeric", "numeric", "text", "numeric", "numeric", "numeric", "numeric", "numeric", "date", "date", "numeric", "numeric", "date", "date",  "numeric", "date", "date", "numeric", "text", "numeric", "text", "numeric", "numeric", "date", "numeric", "numeric", "text", "numeric", "text", "text"))
#Load imaging/score data
images <- read_excel("~/Documents/ROCIanalysis/ROCI_101918.xlsx", sheet = "Imaging", col_types = c("numeric", "text", "date", "numeric", "date", "numeric", "numeric", "numeric", "date", "text", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "skip"))
```

<h3>Analysis of the non-control cohort who have a CXR from MICU admit (n=559)</h3>
**TLDR**:
<br>Total CXR score is predictive in a threshold fashion. Once separated into the lowest quartile and the upper three quartiles, it is no longer predictive (of in-hospital mortality).
<br> - Total CXR score IS predictive of in-hospital mortality, even when adjusted for APACHE score and immunosuppression
<br> - Total CXR score IS predictive of 28-day mortality, even when adjusted for APACHE score and immunosuppression
<br> - Total CXR score IS predictive of 60-day mortality, even when adjusted for APACHE score and immunosuppression
<br> - Total CXR score IS predictive of time-to-death ina Cox proportional hazards model
<br> - Total CXR score IS predictive of ICU LOS and duration of MV
<br>

```{r filter data}
#Creates one dataset of outcomes and scores for subjects with a CXR from MICU admit
admit <- filter(images, Final_Dx != "Control", admit_cxr == 1) %>% dplyr::select(Subject_ID, cxr_date, Quad1, Quad2, Quad3, Quad4, Total)
imgs <- filter(outcomes, Final_Dx != "Control", imgs == 1) %>% dplyr::select (Subject_ID, Male, Age, BMI, non_white, Immsupp, APACHE, SOFA, days_MV, ICU_admit, ICU_days, Hosp_admit, Hosp_days, Final_Dx, DC_coded, hosp_death, date_death, mort28d, mort60d) %>% merge(admit, by="Subject_ID")

# 207/559 subjects are immunosuppressed
# 286/559 subjects have a BMI recorded
# 285/559 subjects were intubated
```
```{r more set-up, echo=FALSE}
#Numerics to factor
imgs$Male <- as.factor(imgs$Male)
imgs$non_white <- as.factor(imgs$non_white)
imgs$Immsupp <- as.factor(imgs$Immsupp)
imgs$hosp_death <- as.factor(imgs$hosp_death)
imgs$mort28d <- as.factor(imgs$mort28d)
imgs$mort60d <- as.factor(imgs$mort60d)
```

<br>**In-hospital mortality**: With the exception of age and gender, all of the covariates are significant predictors of in hospital death and none of them appear to be colinear or confounding. Notably, total CXR score remains significant even when adjusting for APACHE score and immunosuppression status. 
```{r logit in-hospital mortality, warning=FALSE}
#In-hospital mortality
hosp_mort <- glm(hosp_death ~ Total+Male+Age+Immsupp+APACHE+non_white, family=binomial(link='logit'), data=imgs)
summary(hosp_mort)

cbind(OR = exp(coef(hosp_mort)),
      exp(confint(hosp_mort)),
      p = coef(summary(hosp_mort))[,4])

```

<br>**28-day mortality**: Age and gender are no longer predictive, but Total CXR score remains significant even when Immunosuppression and APACHE are adjusted for. 

```{r logit 28d mortality}
one_month <- glm(mort28d ~ Total+Male+Age+Immsupp+APACHE+non_white, family=binomial(link='logit'), data=imgs)
summary(one_month)

cbind(OR = exp(coef(one_month)),
      exp(confint(one_month)),
      p = coef(summary(one_month))[,4])
```
<br>**60-day mortality**: Age and gender are no longer predictive, but Total CXR score remains significant even when Immunosuppression and APACHE are adjusted for. 

```{r logit 60d mortality}
two_month <- glm(mort60d ~ Total+Male+Age+Immsupp+APACHE+non_white, family=binomial(link='logit'), data=imgs)
summary(two_month)

cbind(OR = exp(coef(two_month)),
      exp(confint(two_month)),
      p = coef(summary(two_month))[,4])
```
<br>**Time to death/Cox model**:  The total CXR score is the only predictive covariate in a cox proportional hazards model. 
```{r cox proportional hazards}
refdate <- as.Date("2018-09-15")
#Create variables with survival (in days) and a binary indicating censored status
imgs <- mutate(imgs, survival = ifelse(is.na(date_death) == TRUE, as.numeric(difftime(refdate, ICU_admit, units='days')), as.numeric(difftime(date_death, ICU_admit, units='days')))) %>% mutate(censor = ifelse(is.na(date_death) == TRUE, 0, 1))

#Cox regression model
cox_all <- coxph(Surv(survival, censor) ~ Male+Age+non_white+APACHE+Immsupp+Total, data=imgs)
summary(cox_all)

#non-ARDS patients
imgs_non <- imgs %>% filter(Final_Dx != "ARDS" & Final_Dx != "Sepsis/ARDS")
cox_all_non <- coxph(Surv(survival, censor) ~ Male+Age+non_white+APACHE+Immsupp+Total, data=imgs_non)
summary(cox_all_non)
```
**Cox by quartile**: 
```{r quartiles}
#create quartiles
imgs <- imgs %>% mutate(quartile = ntile(Total,4)) 
imgs$quartile <- as.factor(imgs$quartile)
#Create factor of upper three quartiles (vs lowest)
imgs <- imgs %>% mutate( upthree = ifelse(quartile == 1, 0, 1))
imgs$upthree = as.factor(imgs$upthree)
#Cox regressions
cox_qt <- coxph(Surv(survival, censor) ~ Age+Male+Immsupp+non_white+APACHE+strata(quartile), data=imgs) #for graph
cox_quart <- coxph(Surv(survival, censor) ~ Age+Male+Immsupp+non_white+APACHE+upthree, data=imgs) #for betas
summary(cox_quart)
```
```{r Kaplan Meier}
ggsurvplot(survfit(cox_qt), data=imgs, 
           ggtheme = theme_minimal(), legend.title="Quartile",
           risk.table=TRUE, tables.height = 0.2, tables.theme = theme_cleantable(),
           xlim=c(0,3650), break.x.by=365,
           title="Survival by CXR score quartile", xlab="Time (days)",
           legend="right",
           risk.table.fontsize=3, risk.table.title="") 
  
```
```{r in-hosp mort by quartile}
hosp_mort <- glm(hosp_death ~ upthree+Male+Age+Immsupp+APACHE+non_white, family=binomial(link='logit'), data=imgs)
summary(hosp_mort)

cbind(OR = exp(coef(hosp_mort)),
      exp(confint(hosp_mort)),
      p = coef(summary(hosp_mort))[,4])
```

<br>**Other outcome measures**: Total is predictive of ICU LOS and duration of MV 
```{r ICU LOS, warning=FALSE}
contrasts(imgs$quartile) <-contr.treatment(4,base = 1)

#Negative binomial model
ICUdays_nb <- glm.nb(ICU_days ~ Total+APACHE+Age+non_white+Male+Immsupp, data=imgs)
summary(ICUdays_nb)
cbind(IRR = exp(coef(ICUdays_nb)), exp(confint(ICUdays_nb)))[2,]

ICUdays_nbq <- glm.nb(ICU_days ~ quartile+APACHE+Age+non_white+Male+Immsupp, data=imgs)
summary(ICUdays_nbq)
cbind(IRR = exp(coef(ICUdays_nbq)), exp(confint(ICUdays_nbq)))[2:4,]
```

```{r mechanical ventilation, warning=FALSE}

#Negative binomial model
MVdays_nb <- glm.nb(days_MV ~ Total+APACHE+Age+non_white+Male+Immsupp, data=imgs)
summary(MVdays_nb)
cbind(IRR = exp(coef(MVdays_nb)), exp(confint(MVdays_nb)))[2,]

MVdays_nbq <- glm.nb(days_MV ~ quartile+APACHE+Age+non_white+Male+Immsupp, data=imgs)
summary(MVdays_nbq)
cbind(IRR = exp(coef(MVdays_nbq)), exp(confint(MVdays_nbq)))[2:4,]
```

```{r boxplot}
library(reshape2)
library(RColorBrewer)
#Eliminate non-intubated from the MV data and transform for plotting
meltICU <- imgs %>% dplyr::select(Subject_ID, ICU_days, quartile) %>% melt(id.vars="quartile", measure.vars="ICU_days")
meltMV <- imgs %>% filter(days_MV != 0) %>% dplyr::select(Subject_ID, days_MV, quartile) %>% melt(id.vars="quartile", measure.vars="days_MV")
meltplot <- rbind(meltICU, meltMV)

#Generate the boxplot
ggplot(meltplot) + geom_boxplot(aes(quartile, value, color=variable, fill=quartile, line=quartile)) +   
  scale_x_discrete(name="CXR Score Quartile") + 
  scale_y_sqrt(name="Days", breaks=c(0,4,9,16,25,36,49,64,81)) +
  ggtitle("Duration of Mechanical Ventilation & ICU Length of Stay") +
  theme_minimal() + theme(plot.title = element_text(face = "bold")) +
  theme(panel.grid.minor = element_blank()) + 
  guides(color=guide_legend(title = "Outcome"))

```


```{r zero-inflated negative binomial, warning=FALSE}
#good for count data with lots of zeros (eg duration of MV)
#library(pscl)
#library(MASS)
#library(boot)

#look at the data - we see that zero is over-represented and the data are highly non-normal (bad for OLS) with a wide dispersion (bad for zero-inflated Poisson).
#ggplot(aes(days_MV), data=imgs) + geom_histogram(binwidth = 2)

#zinb <- zeroinfl(days_MV ~ Total | non_white+Male+APACHE+Immsupp, data=imgs, dist="negbin", EM=TRUE)
#summary(zinb)
```


```{r per point for quartile subgroups}
quart1 <- imgs %>% filter(upthree == 0)
quart234 <- imgs %>% filter(upthree == 1)

hosp_mort <- glm(hosp_death ~ Total+Male+Age+APACHE+Immsupp+non_white, family=binomial(link='logit'), data=quart1)
summary(hosp_mort)

cbind(OR = exp(coef(hosp_mort)),
      exp(confint(hosp_mort)),
      p = coef(summary(hosp_mort))[,4])

hosp_mort <- glm(hosp_death ~ Total+Male+Age+Immsupp+non_white, family=binomial(link='logit'), data=quart234)
summary(hosp_mort)

cbind(OR = exp(coef(hosp_mort)),
      exp(confint(hosp_mort)),
      p = coef(summary(hosp_mort))[,4])
```


```{r in-hospital mortality: subgroup by immunosuppression}
competent <- imgs %>% filter(Immsupp == 0)
compromised <- imgs %>% filter(Immsupp == 1)

hosp_mort <- glm(hosp_death ~ Total+Male+Age+APACHE+non_white, family=binomial(link='logit'), data=competent)
summary(hosp_mort)

cbind(OR = exp(coef(hosp_mort)),
      exp(confint(hosp_mort)),
      p = coef(summary(hosp_mort))[,4])

hosp_mort <- glm(hosp_death ~ Total+Male+Age+APACHE+non_white, family=binomial(link='logit'), data=compromised)
summary(hosp_mort)

cbind(OR = exp(coef(hosp_mort)),
      exp(confint(hosp_mort)),
      p = coef(summary(hosp_mort))[,4])
```

```{r cox survival: subgroup by immunosuppression}
cox_all <- coxph(Surv(survival, censor) ~ Male+Age+non_white+APACHE+Total, data=competent)
summary(cox_all)

cox_all <- coxph(Surv(survival, censor) ~ Male+Age+non_white+APACHE+Total, data=compromised)
summary(cox_all)
```



