---
title: "CXR scores and clinical outcomes"
author: "Stefanie Mason"
output: 
  html_document: 
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
---
```{r setup, echo=FALSE, message=FALSE}
library(readxl)
library(tidyverse)
library(survival)
library(survminer)
library(QuantPsyc)
library(gridExtra)
library(irr)
library(reshape2)
library(RColorBrewer)
library(olsrr)
library(broom)
library(data.table)
```
## Disclaimer
The data used in this project are protected by an IRB at Brigham and Women's Hospital. All PHI have been removed from the dataset that is provided for grading. Where applicable, I have described work I could not reveal in order to create this de-identified data. Also, please note that this work has been submitted to a national meeting and thus these results are embargoed from public dissemination. For that reason, my website and screencast are not currently publically viewable. 

# Introduction
## Background
Over 25% of hospital stays in A given year involve some interaction with the intensive care unit (ICU). The ICU is where the sickest patients are cared for: it has the highest nursing ratios, specially trained physicians, and is capable of advanced care such as breathing machines (mechanical ventilation) and even extracorporeal membrane oxygenation (ECMO). At a tertiary care hospital such as the Brigham, the mortality rate in the ICU is over 40%. 

<p>When patients arrive in the ICU, one of the key questions in the patient's mind (also the family), is whether the medical problem is reversible? Eventually, all patients in the ICU will go down one of two paths: they will improve and leave the ICU or they will die in the ICU. As a clinician in the ICU, answering this question is not only important for counseling a patient (and family), but also for allocating scarce resouces.

<p>So what objective data exists to help a clinican determine how likely a patient is to die during their admission? Currently, there are scoring systems such as the APACHE and the SOFA. These scores require extensive blood work and knowledge of the patient's prior history - lots of data and up to 24 hours worth of time. There are more efficient scoring systems, but they require knowing exactly what the underlying diagnosis is (and assume there is only one), which is frequently not immediately known. Because of these difficulties, scoring systems are applied to only about 15% of patients in actual practice.

## Hypothesis
This study hypothesizes that a clinician can use a chest x-ray (CXR) to predict the risk of death in an ICU patient. CXR are routinely done on almost all ICU patients and are available for review immediately. One prior study (Warren et al 2017) has looked at this idea specifically within the diagnosis of Acute Respiratory Distress Syndrome (ARDS), but this study attempts to generalise the idea to the broader population of all ICU patients, regardless of underlying diagnosis.

To ensure the score was correlating with a physiologic process (and not some form of artifact), I also compared CXRs taken within 24 hours of death with the lung weights at autopsy. Heavier lungs would be expected to be heavier, from blood, inflammation, or edema. This should, in turn, show up on a CXR as opacification of the airspaces (which increases the CXR score).

Lastly, I looked for an association between the CXR score and some biomarkers that have been linked to sepsis and ICU mortality.

# Methods
## Study population
The data used for this study comes from the Registry of Critical Illness (ROCI) at Brigham and Women's hospital. The registry is a convenience sample of patient consented within 24 hours of admission to the ICU between August 2008 and the present day. For the purposes of this analysis, I closed the cohort August 2018 (n= 650) and excluded patients whom did not have a CXR within 24 hours of arrival to the ICU (n= 65). I also exluded patients that were admitted for an observation-only protocol, such as desensitization to an antibiotic (n = 25).

## CXR scoring
<center><iframe src="https://drive.google.com/file/d/1AFAdPmfxjPzPIVHpX2PJbIxBuuBS2hcn/preview" width="640" height="480"></iframe></center>

<p>To calculate a visual score for each CXR, the image was divided into four quadrants. Each quadrant was assigned a score from 0 to 4 with:
<br> 0 = normal
<br> 1 = hazy opacification of less than half the quadrant
<br> 2 = dense opacification of less than half the quadrant
<br> 3= hazy opacification of more than half the quadrant
<br> 4 = dense opacification of more than half the quadrant. 
<p> The total score, ranging from 0 to 16, was the sum of all four quadrants. 

# Analysis
## Wrangling
The following wrangling occurred prior to de-identification of the data:
<br> - A pull of demographic data from the EMR was imported and left joined to a list of study medical record numbers to gather gender, race, and birth date data (by medical record number)
<br> - The difference between birth date and hospital admission was used to calculate age at index admission
<br> - After manually reviewing the charts, the difference between ICU admit and discharge dates was used to calculate ICU days. A similar calculation was done for hospital length of stay.
<br> - The survminor package was used to calculate survival in days as well as noting which subjects were censored
<br> - The ICU admission date was compared to the list of reviewed images. Any chest x-ray within 24 hours was flagged as an admit image. Any subject without an admit CXR was flagged for manual chart review. 
<br> - A sample of 50 chest x-rays was pulled for inter-rater reliability review (had to include medical record number to locate the image)

## Inter-rater reliability
```{r import IRR file}
IRR <- read_excel("ROCI_clean.xlsx", sheet = "IRR")
#Reorganise data from rows into columns
IRR <- IRR %>% dplyr::select(-Subject_ID, -Quad1, -Quad2, -Quad3, -Quad4) %>% unite(img_date, CXR_date, CXR_time, sep="_") %>% spread(Reviewer, Total) 
names(IRR) <- c("img_id", "Rev1", "Rev2")
```

<p>Calculation of a two-way, mixed consistency single-measures intraclass correlation coefficient:
```{r ICC calculation}
icc(IRR[c(2,3)], model="t", type="c", r0=0, conf.level=0.95)
```

<p>Our coefficient demonstrates good agreement between raters. Let's look at this visually with a simple scatterplot as well as a Bland-Altman plot.

```{r scatterplot}
scatter <- ggplot(IRR, aes(Rev1, Rev2)) + geom_point(size=2) + 
  ggtitle("Agreement between Reviewers") + xlab("Reviewer 1 score") + ylab("Reviewer 2 score") + 
  geom_abline(slope=1, size=0.5, color="blue") + 
  scale_x_continuous(breaks=seq(0,16,2)) + scale_y_continuous(breaks=seq(0,16,2))
```

```{r BA plot}
#Generate the differences and averages for plot
IRR <- mutate(IRR, diff = (Rev1 - Rev2), avg = ((Rev1 + Rev2) / 2))
#Generate plot
BAplot <- ggplot(IRR, aes(avg, diff)) + geom_point(size=2) + 
  ggtitle("Bland-Altman Plot") + xlab("Average") + ylab("Difference") +
  geom_hline(yintercept=mean(IRR$diff), color="blue", size=0.5) + 
  geom_hline(yintercept=mean(IRR$diff) - 1.96*sd(IRR$diff), color="red", size=0.5) + 
  geom_hline(yintercept=mean(IRR$diff) + 1.96*sd(IRR$diff), color="red", size=0.5)
```

```{r plot together}
grid.arrange(scatter, BAplot, ncol=2)
```

## Mortality

The following modifications need to be made to the data:
<br> - Exclude people coded as "Control"
<br> - Race re-coded to a binary variable (white, non-white)
<br> - A binary factor for in-hospital death created based on discharge location
<br> - Columns imported as numeric re-coded to factors (Male, Immunosuppression)
<br> - Join it with the image score data
<br> - Exclude anyone without a CXR within 24H of admission

```{r import regression data, message=FALSE}
#Load data
outcomes <- read_excel("ROCI_clean.xlsx", sheet = "Outcomes", range = cell_cols(1:17))
images <- read_excel("ROCI_clean.xlsx", sheet = "Imaging", range = cell_cols(1:9))

#RWarngle data
outcomes <- outcomes %>% mutate(non_white = as.factor(ifelse(Race == "White, non-Hispanic origin", 0, 1)), hosp_mort = as.factor(ifelse(DC_coded == 0 | DC_coded == 1, 1, 0)), Immsupp = as.factor(Immsupp), Male = as.factor(Male))

images <- images %>% dplyr::select(-day_3_cxr, -day_5_cxr) %>% filter(admit_cxr == 1) %>% left_join(outcomes, by = "Subject_ID") %>% filter(Final_Dx != "Control") 

#This results in a data frame with 560 observations, which is what we expect
```

### In-hospital mortality
For in-hospital mortality, our outcome is a binary state: you survived to discharge or you did not. This suggests we should use a logistic regression. It is common in the ICU literature to control for age, race, and gender. I will also control for immunosuppression and their illness severity (as calculated by the APACHE score).

```{r logit in-hospital mortality, message=FALSE}
#In-hospital mortality
inhosp <- glm(hosp_mort ~ Total+Male+Age+Immsupp+APACHE+non_white, family=binomial(link='logit'), data=images)
summary(inhosp)

cbind(OR = exp(coef(inhosp)),
      exp(confint(inhosp)),
      p = coef(summary(inhosp))[,4])
```
<p> This data indicates that the CXR score (variable = Total) is a significant predictor of in-hospital death. It suggests that for every 1-point increase in total CXR score, there is a 9.47% increased risk of death (CI: 4.34-15.0).

<p> But a previous paper by Warren et. al. suggested noted a significant difference between quartiles 1&2 compared to 3&4. This suggests there may be a threshold effect to survival. So is the CXR score equally predictive across quartiles? 

<p> First, let's take a quick look at what range of CXR scores fall in each quartile.
```{r quartile}
#Make quartiles
images <- images %>% mutate(quartile = as.factor(ntile(Total,4)))

#Summarize quartiles
images %>% group_by(quartile) %>% summarise(min_score = min(Total), max_score = max(Total), median = median(Total), n = length(Total))
```

<p>Now, let's re-run the regression with quartiles. 
```{r logit by quartile, message=FALSE}
#Re-run regression
inhosp_quart <- glm(hosp_mort ~ quartile+Male+Age+Immsupp+APACHE+non_white, family=binomial(link='logit'), data=images)
summary(inhosp_quart)

cbind(OR = exp(coef(inhosp_quart)),
      exp(confint(inhosp_quart)),
      p = coef(summary(inhosp_quart))[,4])
```
<p> The result is significant across all quartiles (with lower scores = quartile 1 being the reference group) with increasing OR for each successive quartiles. Similar to the previous paper, however, the difference between quartiles 3 and 4 is not large. The second quartile has 2.7 times the odds of in-hospital death, the third quartile 3.3 times the odds, and the highest quartile has 3.5 times the odds of in-hospital death compared with the lowest quartile. 

### Overall mortality
So now that we know the CXR score predicts in-hospital mortality, does it also predict overall mortality? For this, I will use a cox regression. Again, to better explore the nature of the association, this is done in quartiles. The result will be visualized with a Kaplan Meier curve.

```{r cox proportional hazards}
#Cox regression model
cox_quart <- coxph(Surv(Survival, censor) ~ Age+Male+Immsupp+non_white+APACHE+strata(quartile), data=images) #for KM
cox_qt <- coxph(Surv(Survival, censor) ~ Age+Male+Immsupp+non_white+APACHE+quartile, data=images) #for HR
summary(cox_qt)
```

```{r Kaplan Meier all subjects}
KM_all <- ggsurvplot(survfit(cox_quart), data=images, 
           ggtheme = theme_minimal(), legend.title="Quartile", legend.labs=c("Q1","Q2","Q3","Q4"),
           risk.table=TRUE, tables.height = 0.2, tables.theme = theme_cleantable(),
           xlim=c(0,3650), break.time.by=365.25, xscale="d_m", ylab="",
           xlab="Months", font.legend=8, title="Survival by CXR score quartile", font.title=10,
           legend=c(1.1,0.6), risk.table.y.text=FALSE,
           risk.table.fontsize=3, risk.table.title="") 
KM_all$plot <- KM_all$plot + theme(plot.title = element_text(hjust = 0.5))
KM_all
```

<p> The overall survival data suggests more of a threshold effect than the in-hospital death data did. The hazard ratios for each quartile are significant, but are fairly similar for quartiles 2, 3, and 4 compared to the lowest quartile. This is most striking when looking at the KM plot: the quartile 1 curve (orange) is clearly separated from the other three. 

<p>Could all of this be driven by the subset of patients in this cohort with acute respiratory distress syndrome (ARDS)? In other words, does the relationship remain true when we remove the sickest respiratory patients and instead look mostly at people with blood infections/inflammation (what we call sepsis)? To find out, let's re-run the Cox and re-draw the KM without the ARDS patients (n=122).

```{r Cox regression for non ARDS}
#non-ARDS patients
imgs_non <- images %>% filter(Final_Dx != "ARDS" & Final_Dx != "Sepsis/ARDS") %>% mutate(quartile = as.factor(ntile(Total,4)))
cox_non <- coxph(Surv(Survival, censor) ~ Male+Age+non_white+APACHE+Immsupp+strata(quartile), data=imgs_non)  #for KM
cox_non_hr <- coxph(Surv(Survival, censor) ~ Male+Age+non_white+APACHE+Immsupp+quartile, data=imgs_non) 
summary(cox_non_hr)
```

```{r KM for non ARDS}
KM_non <- ggsurvplot(survfit(cox_non), data=imgs_non, 
           ggtheme = theme_minimal(), legend.title="Quartile", 
           risk.table=TRUE, tables.height = 0.2, tables.theme = theme_cleantable(),
           xlim=c(0,3650), break.time.by=365.25, xscale="d_m", ylab="",
           xlab="Months", font.legend=8, title="Survival by CXR score quartile in non-ARDS patients", font.title=10,
           legend="none", risk.table.y.text=FALSE, 
           risk.table.fontsize=3, risk.table.title="") 
KM_non$plot <- KM_non$plot + theme(axis.text.y = element_blank(), plot.title = element_text(hjust = 0.5))
KM_non
```

<p> The relationship between the CXR score and survivial is essentially identical despite the removal of the ARDS patients. Again we see a threshold effect between the lowest quartile and the upper three. What is of particular interest about this threshold, is that it is quite low. Recall the quartile ranges described above. The lowest quartile were scores from 0-2. This suggests that even a mild abnormality on the CXR has significant implications for prognosis in ICU patients. The previously mentioned Warren et al paper also demonstrated an association between CXR score and survival, however their association was no longer significant when controlling for APACHE score. It is possible that given their cohort was entirely made of ARDS patients, all of their subjects fell above quartile 1 and thus they were unable to discern this very low threshold effect. 

## Other Outcome Measures
### Length of ICU stay

<p>Let's turn away from mortality and look at some other clinical measures such as length of stay in the ICU and duration of mechanical ventilation.

```{r ICU LOS, warning=FALSE, message=FALSE}
contrasts(images$quartile) <-contr.treatment(4,base = 1)

#Negative binomial model
ICUdays_nb <- glm.nb(ICU_days ~ Total+APACHE+Age+non_white+Male+Immsupp, data=images)
summary(ICUdays_nb)
cbind(IRR = exp(coef(ICUdays_nb)), exp(confint(ICUdays_nb)))[2,]
```

```{r mechanical ventilation, warning=FALSE, message=FALSE}
#Negative binomial: intub patients only
intub <- images %>% filter(days_MV != 0)
MVdays_nb <- glm.nb(days_MV ~ Total+APACHE+Age+non_white+Male+Immsupp, data=intub)
summary(MVdays_nb)
cbind(IRR = exp(coef(MVdays_nb)), exp(confint(MVdays_nb)))[2,]
```

```{r boxplot}
#Eliminate non-intubated from the MV data and transform for plotting
meltICU <- images %>% dplyr::select(Subject_ID, ICU_days, quartile) %>% melt(id.vars="quartile", measure.vars="ICU_days")
meltMV <- intub %>% dplyr::select(Subject_ID, days_MV, quartile) %>% melt(id.vars="quartile", measure.vars="days_MV")
meltplot <- rbind(meltICU, meltMV)

#Generate the boxplot
ggplot(meltplot) + geom_boxplot(aes(quartile, value, color=variable)) +   
  scale_x_discrete(name="CXR Score Quartile") + 
  scale_y_sqrt(name="Days", breaks=c(0,4,9,16,25,36,49,64,81)) +
  ggtitle("Duration of Mechanical Ventilation & ICU Length of Stay") +
  theme_minimal() + theme(plot.title = element_text(face = "bold")) +
  theme(panel.grid.minor = element_blank()) + 
  guides(color=guide_legend(title = "Outcome"))

```

<p> These data demonstrate that for each one point increase in CXR score, ICU length of stay increased by 5.72% (Rate ratio 1.06, CI 1.04-1.07). For each one point increase in CXR, the duration of mechanical ventilation (for intubated patients) increased 4.44% (rate ratio 1.04, CI 1.02-1.07). 

## Lung weights

<p>To ensure the CXR score is measuring a physiologic state rather than an artifact of imaging, the CXR scores were compared to lung weights measured at autopsy. For the patients in the cohort whom died in the hospital and their family consented to autopsy, the last CXR prior to death (must be within 24 hours) was scored. Since each lung is weighted seperately the weight of the right lung was compared to the sum of quartiles 1&3 (the right sided quartiles) and the weight of the left lung was compared to the sum of quartiles 2&4 (the left sided quartiles). 

```{r load lung weight data}
#Load imaging/weight data and merge
weights <- read_excel("ROCI_clean.xlsx", sheet = "LungWt", range = cell_cols(1:10)) %>% 
  gather(side, weight, r_lung_wt:l_lung_wt) %>%
  mutate(side = ifelse(side == "r_lung_wt", "right", "left"), side_total = ifelse(side == "right", quad1+quad3, quad2+quad4)) %>%
  filter(wi_24h == 1) 

weights <- outcomes %>% dplyr::select(Subject_ID, Male) %>% right_join(weights, by="Subject_ID")
```

### Normality
<p>Shapiro Wilk testing was used to test the normality of the distribution of lung weights to be analysed. The distribution was visually inspected with a histrogram.

```{r normality test, warning=FALSE}
shapiro.test(weights$weight)
weights %>% ggplot(aes(weight)) + geom_histogram(bins = 25)
```

<p>The data are non-normally distributed. Warren et al correlated their CXR score with lung weights at explant in lung transplant donors. They used a Spearman correlation. However, this would not allow me to correct for height, which is an important predictor of lung size. Since I want to account for height, I will instead use a linear regression (continuous predictor of a continuous outcome). I will plan to assess the normality of residuals to be sure the model is appropriate. 

```{r linear regression, warning=FALSE}
weights %>% ggplot(aes(side_total, weight, color = side)) + geom_point() + geom_smooth(method = "lm") + xlab("CXR score") +
  ylab("Weight (g)") + ggtitle("CXR score as a predictor of lung weight at autopsy")

wt <- lm(weight ~ side_total+height+Male, data = weights)
summary(wt)
```

<p>Even accounting for height and gender, the CXR score is a significant predictor of lung weight at autopsy. The supports the concept that the opacities seen (and scored) on the CXR correlate to a physiologic process (evident at autopsy).

<p> Since the data were non-normal, let's check a QQ plot and see if the residuals are normally distributed.

```{r qq plot}
#check normality of errors
ols_plot_resid_qq(wt)
ols_test_correlation(wt)  #expected vs predicted residuals
```

<p>The residual testing suggests a linear regression model was reasonable despite the non-normality of the input data.

## Biomarkers

<p>Researchers have long been looking for something they can measure in blood work that will predict clinical outcomes. No single measure has proven definitive, which is one reason the scoring systems use so many different measures. For example, procalcitonin is a marker of bacterial infection, but it is not elevated in viral or fungal disease (or non-infectious disease). In contrast, mitchondrial DNA (mtDNA) has been associated with in-ICU death and, in patients with levels >3,200, with overall survival time, regardless of underlying diagnosis. One of the drawbacks is that the association does not hold in cardiac ICU or neurologic ICU patients. But this dataset is medical ICU patients, so I would expect the CXR score to be associated with mtDNA levels.

```{r import biomarker data}
#Biomarker data
biomarker <- read_excel("ROCI_clean.xlsx", sheet="markers", range = cell_cols(1:9))
biomarker <- biomarker %>% dplyr::select(-UO_24) %>% rename(Subject_ID = SID) 
biomarker <- images %>% dplyr::select(Subject_ID, Total, APACHE) %>% right_join(biomarker, by="Subject_ID")
```

###Data summary

<p>Let's take a look at how much biomarker data exists and it looks like.

```{r summary stats}
#This isn't the most efficient way of accomplishing this, but I wanted to demonstrate functions and apply()
markers <- biomarker %>% dplyr::select(IL18D0, nadh_copies, HMITO_38, HMITO_96, bglob, Lactate)

#define functions to accomodate missing data
count_rows <- function(x){
  NROW(na.omit(x))
}
max_val <- function(x){
  max(na.omit(x))
}
min_val <- function(x){
  min(na.omit(x))
}

#Build the summary table
sum_stat <- data.table(IL18D0=double(), nadh_copies=double(), HMITO_38=double(), HMITO_96=double(), bglob=double(), Lactate=double())
sum_stat <- rbind(sum_stat, t(apply(markers, 2, count_rows)))
sum_stat <- rbind(sum_stat, t(apply(markers, 2, max_val)))
sum_stat <- rbind(sum_stat, t(apply(markers, 2, min_val)))
sum_stat <- cbind(sum_stat, c("n", "max", "min"))
sum_stat <- sum_stat[,c(7,1,2,3,4,5,6)]
sum_stat
```

```{r biomarker scatterplot, warning=FALSE}
biomarker %>% dplyr::select(IL18D0, nadh_copies, HMITO_38, HMITO_96, bglob, Lactate, Total) %>% gather(marker, value, -Total, na.rm=T) %>% 
  ggplot(aes(Total, value)) + geom_point() + scale_y_log10() + facet_grid(.~marker) + ylab("value (log scale)") + xlab("Total CXR score")
```

<p>Truthfully, I am not going to use all these biomarkers. The literature is strongest for IL-18 (inflammatory marker), nadh (which represents mitchondrial DNA), and bglob (which is nuclear DNA). So going forward, those are the ones I will work with.

###Normality
<p>Some of those distributions do not look normal. Let's check them with the Shapiro-Wilk test and then see if transformations would help.

```{r normality}
#Original scale
shapiro.test(biomarker$IL18D0)
shapiro.test(biomarker$nadh_copies)
shapiro.test(biomarker$bglob)

#Log transforms
biomarker <- biomarker %>% mutate(logIL18 = log10(IL18D0), logbglob = log10(bglob), lognadh = log10(nadh_copies))

shapiro.test(biomarker$logIL18)
shapiro.test(biomarker$lognadh)
shapiro.test(biomarker$logbglob)
```

<p>NADH and IL-18 log transform to normal. Bglob does not, but it's closer. 

###Univariable model

<p>Let's see if the CXR score predicts the blood level of the biomarker.

```{r univariate regression}
biomarker %>% dplyr::select(logIL18, lognadh, logbglob, Total) %>% gather(marker, value, -Total, na.rm=T) %>%
  group_by(marker) %>% do(tidy(lm(value~Total, data = .))) %>% filter(term != "(Intercept)")

```

```{r scatterplot with log transforms, warning=FALSE}
biomarker %>% dplyr::select(logIL18, lognadh, logbglob, Total) %>% gather(marker, value, -Total, na.rm=T) %>% 
  ggplot(aes(Total, value)) + geom_point() + facet_grid(.~marker) + ylab("value (log scale)") + xlab("Total CXR score") + geom_smooth(method = "lm")
```

<p>In an unadjusted analysis, the CXR score is significantly associated with mtDNA, nucDNA, but not the IL-18 inflammatory marker. What happens when I control for APACHE score?

###Multivariable model

```{r multivariate regression}
biomarker %>% dplyr::select(logIL18, lognadh, logbglob, Total, APACHE) %>% gather(marker, value, -Total, -APACHE, na.rm=T) %>%
  group_by(marker) %>% do(tidy(lm(value~Total+APACHE, data = .))) %>% filter(term != "(Intercept)")
```

<p>Unfortunately the associations are no longer significant once I control for the APACHE score. While it would be ideal if the association held, the biomarkers themselves are not perfect proxies for survival. 

## Conclusions
<p>The CXR score is associated with in-hospital mortality, overall survival, ICU length of stay, and duration of mechanical ventilation. Interestingly, the association with overall survival appears to be a threhold effect, with very mild abnormalities on CXR confering a significant increased risk of overall mortality. The score appears to me picking up true physiologic changes, given its association with lung weights at autopsy and serum biomarkers. 

###Next steps
<p> As a follow up to this study, I am hoping to generate a convolutional neural network that will predict survival based on CXR data.


